Generalized Bradley-Terry models

### 2025-03-20 inst/jss_hyper3_real.Rnw
typo

Specifically, it seems reasonable to assume that if we ask our
respondent to rank his options, it is highly probable that he will put
RB and BB consecutively (because they are essentially
indistinguishable).  Can we quantify the strength of this effect?  To
do this, we define a bespoke function \fct{RB\_BB\_LF} which
returns a \code{hyper3} log-likelihood function corresponding to
repeated observations of our commuter's reported ranks for the five
options.
%
<<defineredbuslf>>=
`RB_BB_LF` <- function(lambda){
    ec <- c(C=1, T=2, RB=3, BB=3, W=4) 
    h <- c(1, 1, lambda, 1)           
    (
        cheering3(v=c("RB", "BB", "C" , "T", "W"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "RB", "T" , "C", "W"), e = ec, h = h) * 2 + 
        cheering3(v=c("T" , "BB", "RB", "C", "W"), e = ec, h = h) * 5 + 
        cheering3(v=c("W" , "BB", "RB", "T", "C"), e = ec, h = h) * 4 + 
        cheering3(v=c("C" , "RB", "BB", "W", "T"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "C" , "RB", "T", "W"), e = ec, h = h) * 3
    )
}
@
%
Above, we see from the function body that he reported $RB\succ BB\succ
C\succ T\succ W$ three times [first row], and so on; perhaps his
ranking depends on the weather or how tired he is on any given day.
Vector \code{ec} places the options into equivalence classes;
equivalence class 3 comprises the two buses and the other classes are
unsupported as they are of size one.  Vector \code{h} encodes the
strength of the nonindependence.  Observe that in almost every case he
ranks RB and BB consecutively.  Function \fct{RB_BB_LF} takes
argument \code{lambda} that quantifies the perceived similarity
between RB and BB in the same way as function \fct{oldfirm} above.
%
<<definefunctiono>>=
o <- function(lambda){maxp(RB_BB_LF(lambda), give=1)$`Log-likelihood`} 
@ 
%
Above, we define function \fct{o} which returns the support for a
given value of $\lambda$ at the evaluate [over permissible Bradley
  Terry strengths for the options].  We can treat \fct{o} as a
support function for $\lambda$, and then maximize it:
%
<<label=maxlikesim, cache=TRUE>>=
(osup <- optimize(o, c(10, 40), maximum=TRUE))
@ 
%
So a likelihood ratio test of the null that $\lambda=1$ would be:
%
<<label=LRT_RB, cache=TRUE>>=
(suppdiff <- osup$objective - o(lambda = 1))
@
%
We can thus improve the hypothesis that $\lambda=1$, corresponding to
an unaltered Plackett-Luce system, by over 5.6 units of support by
dint of choosing instead $\lambda = 10.49$.  This easily satisfies
\citeauthor{edwards1992}'s two-units-of-support criterion;
alternatively, we could observe that this observation lies in the tail
region of its asymptotic $\chi^2_1$ distribution (Wilks's theorem),
with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2,df=1,lower.tail=FALSE),3)}.
%
%It is straightforward to find a profile
%support function for $\lambda$, Figure~\ref{profilesupportlambdaRBBB}.
%We see that support of $\log\lambda$ is closely quadratic over a broad
%range.
%
%\begin{figure}[htbp]
%\begin{center}
<<plotproflike, echo=FALSE>>=
if(FALSE){
    lambda <- exp(seq(from=log(0.9), to=log(67), len=19)) 
    L <- sapply(lambda, o)
    plot(log(lambda), L-max(L), type="b")
    abline(h=c(0, -2))
    abline(v=0)
}
@ 
%\caption{A\label{profilesupportlambdaRBBB} profile support function
%  for the red bus, blue bus parameter $\lambda$, synthetic dataset.
%  Vertical line indicates null of $\lambda=1$, horizontal lines show
%  maximum support and $-2$}
%\end{center}
%\end{figure}

>>>>

Specifically, it seems reasonable to assume that if we ask our
respondent to rank his options, it is highly probable that he will put
RB and BB consecutively (because they are essentially
indistinguishable).  Can we quantify the strength of this effect?  To
do this, we define a bespoke function \fct{RB\_BB\_LF} which
returns a \code{hyper3} log-likelihood function corresponding to
repeated observations of our commuter's reported ranks for the five
options.
%
<<defineredbuslf>>=
`RB_BB_LF` <- function(lambda){
    ec <- c(C=1, T=2, RB=3, BB=3, W=4) 
    h <- c(1, 1, lambda, 1)           
    (
        cheering3(v=c("RB", "BB", "C" , "T", "W"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "RB", "T" , "C", "W"), e = ec, h = h) * 2 + 
        cheering3(v=c("T" , "BB", "RB", "C", "W"), e = ec, h = h) * 5 + 
        cheering3(v=c("W" , "BB", "RB", "T", "C"), e = ec, h = h) * 4 + 
        cheering3(v=c("C" , "RB", "BB", "W", "T"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "C" , "RB", "T", "W"), e = ec, h = h) * 3
    )
}
@
%
Above, we see from the function body that he reported $RB\succ BB\succ
C\succ T\succ W$ three times [first row], and so on; perhaps his
ranking depends on the weather or how tired he is on any given day.
Vector \code{ec} places the options into equivalence classes;
equivalence class 3 comprises the two buses and the other classes are
unsupported as they are of size one.  Vector \code{h} encodes the
strength of the nonindependence.  Observe that in almost every case he
ranks RB and BB consecutively.  Function \fct{RB\_BB\_LF} takes
argument \code{lambda} that quantifies the perceived similarity
between RB and BB in the same way as function \fct{oldfirm} above.
%
<<definefunctiono>>=
o <- function(lambda){maxp(RB_BB_LF(lambda), give=1)$`Log-likelihood`} 
@ 
%
Above, we define function \fct{o} which returns the support for a
given value of $\lambda$ at the evaluate [over permissible Bradley
  Terry strengths for the options].  We can treat \fct{o} as a
support function for $\lambda$, and then maximize it:
%
<<label=maxlikesim, cache=TRUE>>=
(osup <- optimize(o, c(10, 40), maximum=TRUE))
@ 
%
So a likelihood ratio test of the null that $\lambda=1$ would be:
%
<<label=LRT_RB, cache=TRUE>>=
(suppdiff <- osup$objective - o(lambda = 1))
@
%
We can thus improve the hypothesis that $\lambda=1$, corresponding to
an unaltered Plackett-Luce system, by over 5.6 units of support by
dint of choosing instead $\lambda = 10.49$.  This easily satisfies
\citeauthor{edwards1992}'s two-units-of-support criterion;
alternatively, we could observe that this observation lies in the tail
region of its asymptotic $\chi^2_1$ distribution (Wilks's theorem),
with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2,df=1,lower.tail=FALSE),3)}.
%
%It is straightforward to find a profile
%support function for $\lambda$, Figure~\ref{profilesupportlambdaRBBB}.
%We see that support of $\log\lambda$ is closely quadratic over a broad
%range.
%
%\begin{figure}[htbp]
%\begin{center}
<<plotproflike, echo=FALSE>>=
if(FALSE){
    lambda <- exp(seq(from=log(0.9), to=log(67), len=19)) 
    L <- sapply(lambda, o)
    plot(log(lambda), L-max(L), type="b")
    abline(h=c(0, -2))
    abline(v=0)
}
@ 
%\caption{A\label{profilesupportlambdaRBBB} profile support function
%  for the red bus, blue bus parameter $\lambda$, synthetic dataset.
%  Vertical line indicates null of $\lambda=1$, horizontal lines show
%  maximum support and $-2$}
%\end{center}
%\end{figure}
